---
title: "Project 1: COVID-19 Case Fatality Rate by Race"
author: "Zoë McGinnis, Celvi Lisy, Oscar De La Garza, Anusha Karandikar, & Priscila Morales"
date: 2021-10-19
output:
  github_document:
    toc: true
---

*Purpose*: In this project, we learned how to find useful data and perform a state-level population-weighted analysis of current COVID-19 race trends. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

# Dataset Information

We used The COVID Tracking Project “COVID Racial Data Tracker.” This dataset tracks the race and ethnicity data for COVID-19 cases in US states and territories that report it. 51 out of the 56 states and territories track this data. This dataset contains the variables `date`, `state`, `cases`, `deaths`, `hospitalizations`, and `tests` stratified by race. The different races represented were `AIAN` (American Indian/American Native), `Asian`, `Black`, `Latinx`, `multiracial`, `NHPI` (Native Hawaiian/Pacific Islander), `unknown`, and `White`. In order to find the case fatality rate, we computed the deaths over cases per day for each of the different races represented in the data set. 

As far as the data quality, The COVID Tracking Project cross checks all of its data and sources from state and territory public health authorities and statements from state officials. However, the program itself claims that the data is “incomplete and inconsistent.” The dataset is missing death and/or cases data for the following states and territories: `Guam`, `Hawaii`, `Louisiana`, `Montana`, `New York`, `North Dakota`, `Northern Mariana Islands`, `Puerto Rico`, `Virgin Islands` and `West Virginia`. Additionally, the COVID Tracking Project stopped collecting new data as of March 7, 2021, which should be kept in mind when analyzing the data. 

# Background Information

[COVID-19](https://en.wikipedia.org/wiki/Coronavirus_disease_2019) is the disease caused by the virus SARS-CoV-2. On January 30, 2020, the World Health Organization (WHO) declared the outbreak a “public health emergency of international concern” (PHEIC) [1]. Shortly afterward, it became a global pandemic, leading to huge loss of life and tremendous disruption of society, especially those in minority communities. 

The United States is home to persistent racial disparities in health coverage, chronic health conditions, and mortality. Poor health indicators have consistently trended upward among minority groups relative to non-Hispanic White adults [2]. Similarly, persons from racial and ethnic minority groups are disproportionately affected by COVID-19. Racial and ethnic disparities in COVID-19 case fatality rates are likely driven by a higher risk for exposure to COVID-19 and severe COVID-19 disease (e.g minority groups have a higher prevalence of underlying medical conditions), both of which are influenced by social determinants of health.

Using the COVID Tracking Project data set, we assessed monthly trends in the case fatality rates of COVID-19 among racial and ethnic groups in the United States from January 2020 through March 2021 [3]. 

# The Big Picture
<!-- -------------------------------------------------- -->
We want to study COVID-19 in terms of data: case fatality rate (CFR). The case fatality rate (CFR) is simply the number of confirmed deaths divided by the number of confirmed cases. We're going to do a race group analysis in order to get a view of race trends throughout the pandemic. 

# Leading question
How is race correlated to case fatality rate in COVID cases? 

# Acquire the Data
<!-- -------------------------------------------------- -->

1. Racially disaggregated data from the [COVID Tracking Project](https://covidtracking.com/race)

## Load the data set into the following name `df_race` and tidy it.

```{r covid-data}
## TASK: Load the COVID Tracking data with the tibble name.
race_csv = "./data/CRDT.csv"
df_race <- read.csv(race_csv)
```

<!-- -------------------------------------------------- -->

To get a sense of our tidying task, let's take a glimpse at our data source.

```{r race-data}
glimpse(df_race)
```

Now, let's tidy the data.

```{r tidy}
df_cfr <- df_race %>%
  filter(
    State %in% state.abb
  ) %>%
  mutate(
    State = state.name[match(State, state.abb)],
    Date = as.Date(as.character(Date), tryFormats = c("%Y%m%d"))
  ) %>%
  rename(
    state = State,
    date = Date
  ) %>%
  select(-2) %>% 
  group_by(date) %>%
  summarize(Total_Cases = sum(Cases_Total, na.rm=TRUE),
         Total_White = sum(Cases_White, na.rm=TRUE),
         Total_Black = sum(Cases_Black, na.rm=TRUE),
         Total_Latinx = sum(Cases_Latinx, na.rm=TRUE),
         Total_Asian = sum(Cases_Asian, na.rm=TRUE),
         Total_AIAN = sum(Cases_AIAN, na.rm=TRUE),
         Total_NHPI = sum(Cases_NHPI, na.rm=TRUE),
         Total_Deaths = sum(Deaths_Total, na.rm=TRUE),
         Total_Deaths_White = sum(Deaths_White, na.rm=TRUE),
         Total_Deaths_Black = sum(Deaths_Black, na.rm=TRUE),
         Total_Deaths_Latinx = sum(Deaths_Latinx, na.rm=TRUE),
         Total_Deaths_Asian = sum(Deaths_Asian, na.rm=TRUE),
         Total_Deaths_AIAN = sum(Deaths_AIAN, na.rm=TRUE),
         Total_Deaths_NHPI = sum(Deaths_NHPI, na.rm=TRUE),
         White = (Total_Deaths_White)/Total_White,
         Black = (Total_Deaths_Black)/Total_Black,
         Latinx = (Total_Deaths_Latinx)/Total_Latinx,
         Asian = (Total_Deaths_Asian)/Total_Asian,
         AIAN = (Total_Deaths_AIAN)/Total_AIAN,
         NHPI = (Total_Deaths_NHPI)/Total_NHPI
         ) %>%
  pivot_longer(cols =White: NHPI,
               names_to = "race",
               values_to = "CFR"
               )
```

# Analyzing the data
<!-- -------------------------------------------------- -->
Now that we've done the hard work of loading and wrangling the data, we can finally start our analysis. Our first step will be to produce a graph of deaths/cases, known as the case fatality rate (CFR), over time by race. Then we will explore the data.

## Including Plots

```{r plot}
df_cfr %>%
  ggplot(mapping = aes(x = date, y = CFR, color = race, na.rm = TRUE)) + 
  labs(title = "COVID-19 Case Fatality Rate (CFR) by Race", 
       y = "CFR", 
       x = "Date",
       color = "Race",
       caption = "The y-axis is scaled to more clearly show trendlines; some data exist outside of the plot shown."
       ) +
  theme(axis.title.x = element_text(size = 14, family = "Times New Roman"),
        axis.title.y = element_text(size = 14, family = "Times New Roman"),
        plot.title = element_text(size = 20, family = "Times New Roman", hjust= 0.5),
        legend.title = element_text(size = 16, family = "Times New Roman"),
        legend.text = element_text(size = 12, family = "Times New Roman")
        ) +
  geom_smooth(se=F, na.rm = T) +
  geom_point(na.rm = T, size = .7) +
  ylim(0,.15)
```


# Conclusions 
The graph above shows a gradual decline in case fatality rates across all races besides American Indians and Alaskan Natives (we suspect this may be due to low or inconsistent reporting of deaths on reservations). We speculate that this general downward trend might be a result of improvement in treatment methods, increased awareness of the disease, improvement in public health policies, and reduced strain on hospital capacity and medical personnel.

The data do not appear to suggest any significant disparities in CFR by race; although case fatality rates of Asian and Black populations gradually climb above those of White populations, the trend lines converge toward March of 2021, and do not stray too far from each other throughout the graph. We theorize that if reporting were more consistent, case fatality rates of minorities compared to that of the White population might reflect the racial inequality present in US institutions. 

An interesting feature of the graph is the low CFR of Native Hawaiians and Pacific Islanders. Although the death rate of the population appears to be significantly lower than other races throughout the span of the graph, Hawaii has the largest number of Native Hawaiians and Pacific Islanders [4], and since they do not report data COVID-19 data by race, we suspect this trend may be misleading.

# Remaining Questions
Our next steps would be to use either our existing data or find new data to explore why we saw the patterns we did in the graph. If we choose to focus on the data we have, we could break down the case fatality rate by state to explore which states contribute most to the racial disparities we see in the data.
The data from The COVID Tracking Project includes statistics for both testing and hospitalization rates by race, which could potentially provide insight about the case fatality rates we found. Both testing and hospitalization fall under the umbrella of access to healthcare, which, from our background information research, we believe could be correlated to race. Exploring transmission rates by race (which is not in the data from The COVID Tracking Project) could also give us further information about the relationship between race and environment, and how racial disparities can put people in positions where recommended precautions (ex. social distancing, masking) cannot be fulfilled.

# Works Cited

[1] WHO Europe [WHO](https://www.euro.who.int/en/health-topics/health-emergencies/international-health-regulations/news/news/2020/2/2019-ncov-outbreak-is-an-emergency-of-international-concern) (2020)

[2] PMC article on health disparities [PMC](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3540621/) (2012)

[3] COVID Tracking Tool [COVID Tool](https://covidtracking.com/race) (2021)

[4] The Leadership Conference Fund [NHPI Dataset](http://civilrightsdocs.info/pdf/census/2020/Table1a-States-Number-NHPI-HTC.pdf) (2020)

